Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-15 paper/paper.tex
sketch of the gluing construction for canonicity

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{\hat{\lambda} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
\newcommand{\amp}{\mathrel{\&}}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

\begin{document}
\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a version of dependent
  linear-non-linear type theory in which the tensor product is not
  commutatiev. The linear types can be interpreted as grammars, and
  linear terms as parse transformers. The non-dependent fragment of
  this type theory is already enough to express regular/context-free
  expressions as well as finite automata, with the isomorphism between
  regular languages and automata definable as linear terms. This
  generalizes the language theoretic formalisms such as Kleene
  algebras to incorporate ambiguity. By incorporating a dependency of
  linear types on non-linear types, we can further express indexed
  grammars, pushdown automata and Turing machines. This provides a new
  form of ``logical characterization'' of grammar classes based on a
  substructural logic rather than ordinary first-order logic. Further,
  the type theory can not only express the grammars as types, but
  equivalences between grammars and parsers as terms of the grammar.

@@

\section{Universal Constructions in the Category of Grammars}

Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
and therefore certain initial algebras

\section{A Bunched Type Theory for Grammars}

@@

\subsection{Semantics}

@@

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the semantics just defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``strictly positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu$ and the generators $c$) that every element $t \in
\llbracket A \rrbracket w$ is a kind of tree and that the nodes of the
tree correspond precisely to the introduction forms of the
type. However it is far less obvious that \emph{every} linear term
$\lceil w \rceil \vdash p : \phi$ is equal to some sequence of
introduction forms since proofs can include elimination forms as
well. To show that this is indeed the case we give a \emph{canonicity}
result for the calculus, a generalization of the disjunction theorem
of constructive logic:
\begin{enumerate}
\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item There are no terms $\lceil w \rceil \vdash p : 0$
\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
  to $(M,q)$ where $\cdot \vdash M : A$
\end{enumerate}

@@

\section{Logical Characterizations of Language Classes}

Our bunched type theory

@@

>>>>

\newcommand{\gluedNL}{{\mathcal G}_S}
\newcommand{\gluedNLUniv}{{\mathcal G}_{S,i}}
\newcommand{\gluedL}{{\mathcal G}_L}

\newcommand{\simulsubst}[2]{#1\{#2\}}
\newcommand{\subst}[3]{\simulsubst {#1} {#2/#3}}
\newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\newcommand{\lamb}[2]{\lambda #1.\, #2}
\newcommand{\lamblto}[2]{\lambda^{{\lto}} #1.\, #2}
\newcommand{\lambtol}[2]{\lambda^{{\tol}} #1.\, #2}
\newcommand{\dlamb}[2]{\overline{\lambda} #1.\, #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\applto}[2]{#1 \mathop{{}^{\lto}} #2}
\newcommand{\apptol}[2]{#1 \mathop{{}^{\tol}} #2}
\newcommand{\PiTy}[3]{\Pi #1 : #2.\, #3}
\newcommand{\SigTy}[3]{\Sigma #1 : #2.\, #3}
\newcommand{\LinPiTy}[3]{\widebar\Pi #1 : #2.\, #3}
\newcommand{\LinSigTy}[3]{\widebar\Sigma #1 : #2.\, #3}
\newcommand{\amp}{\mathrel{\&}}
\newcommand{\GrTy}{\mathsf{Gr}}

@@

\begin{document}

\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a version of dependent
  linear-non-linear type theory in which the tensor product is not
  commutative. The linear types can be interpreted as grammars, and
  linear terms as parse transformers. The non-dependent fragment of
  this type theory is already enough to express regular/context-free
  expressions as well as finite automata, with the isomorphism between
  regular languages and automata definable as linear terms. This
  generalizes the language theoretic formalisms such as Kleene
  algebras to incorporate ambiguity. By incorporating a dependency of
  linear types on non-linear types, we can further express indexed
  grammars, pushdown automata and Turing machines. This provides a new
  form of ``logical characterization'' of grammar classes based on a
  substructural logic rather than ordinary first-order logic. Further,
  the type theory can not only express the grammars as types, but
  equivalences between grammars and parsers as terms of the grammar.

@@

%% \section{Universal Constructions in the Category of Grammars}

%% Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
%% and therefore certain initial algebras

\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}

@@

\subsection{Semantics in Grammars}

@@

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$ and the generators $c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

\begin{definition}
  A non-linear type $X$ is purely positive if it is built up using
  only finite sums, finite products and least fixed points.

  A linear type is purely positive if it is built up using only finite
  sums, tensor products, generators $c$, least fixed points and linear
  sigma types over purely positive non-linear types.
\end{definition}

\begin{definition}
  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.

  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
  strings that takes a string $w$ to the definitional equivalence
  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
\end{definition}

\begin{theorem}[Canonicity]
  Let $A$ be a closed, purely positive linear type. Then there is an
  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
\end{theorem}
\begin{proof}
  We outline the proof here, more details are in the appendix. The
  proof proceeds first by a standard logical families construction
  that combines canonicity arguments for dependent type theory
  \cite{coquand,etc} with logical relations constructions for linear
  types \cite{hylandschalk}. It is easy to see by induction that the
  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
  \rrbracket$ and the fundamental lemma proves that the projection
  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
  procedure. Then we establish again by induction that
  canonicalization is also a retraction by observing that introduction
  forms are taken to constructors.
\end{proof}

\begin{enumerate}
\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item There are no terms $\lceil w \rceil \vdash p : 0$
\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
  to $(M,q)$ where $\cdot \vdash M : A$
\end{enumerate}

@@

\section{Type-theoretic Characterizations of Language Classes}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\clearpage
\appendix

\section{Semantics in Sets and Grammars}

A model of Grammar Type Theory is a category with families equipped
with a functor $L : \mathcal C \to \textrm{GrLog}$ from the category
of contexts to the category of grammar logic models and strict morphisms.

First, we give the set theoretic model. We use the standard category
with families structure given by the category of sets and a cumulative
tower of universes.
%
Next, we define $L$ as follows:
\[ L(\Gamma) = \mathcal P \Sigma^*/\textrm{Disc} \Gamma \]
Where $\textrm{Disc} \Gamma$ is the presheaf that is constantly $\Gamma$.
Note that
\[ \mathcal P \Sigma^*/\textrm{Disc} \Gamma \cong (\Set/\Gamma)/\textrm{Disc} \Sigma^*\]
where $\textrm{Disc} \Sigma^*$ is the projection $\Gamma \times \Sigma^*
\to \Gamma$.  Then by the fundamental theorem of topos thoery, $\Set/\Gamma$ is
a topos. Since $\textrm{Disc} : \Set \to \Set/\Gamma$ is monoidal,
$\textrm{Disc} \Sigma^*$ carries a monoid structure in $\Set/\Gamma$ and
this induces a monoidal closed structure on $(\Set/\Gamma)/\textrm{Disc}
\Sigma^*$. Again by the fundamental theorem, it also has finite
products and coproducts and a natural numbers object, and so can
interpret least fixed points.\max{More details about constructing initial algebras?}

Then we can define the remaining type structure as follows. Unraveling
the above, a non-linear type $\Gamma \vdash X$ is interpreted as an
object of $\Set/\sem{\Gamma}$ and a linear type $\Gamma \vdash A$ is
interpreted as an object of $\mathcal P\Sigma^*/\textrm{Disc}\sem{\Gamma}$. We
define them below in indexed style:
\begin{enumerate}
\item $\sem{c} \gamma w = (w = c)$
\item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
\item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
\item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
\item $\sem{L_i}\gamma = (\mathcal P_i\Sigma^*/\textrm{Disc}\sem{\Gamma})_0$
\end{enumerate}

\section{Canonicity}

For the canonicity argument, we construct a second ``glued'' model
that contains a projection morphism to the syntactic model.

Let $S$ be the syntactic category with families for the type
theory. Define the glued category with families to be the comma category
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXG1hdGhjYWwgRyJdLFswLDIsIlMiXSxbMiwyLCJcXFNldCJdLFsyLDAsIlxcU2V0Il0sWzEsMSwiXFxzd2Fycm93Il0sWzMsMl0sWzEsMl0sWzAsMV0sWzAsM11d
\[\begin{tikzcd}[ampersand replacement=\&]
	{\gluedNL} \&\& \Set \\
	\& \rotatebox[origin=c]{-45}{$\Downarrow$} \\
	S \&\& \Set
	\arrow[Rightarrow, no head, from=1-3, to=3-3]
	\arrow["{S(\cdot,-)}"', from=3-1, to=3-3]
	\arrow[from=1-1, to=3-1]
	\arrow[from=1-1, to=1-3]
\end{tikzcd}\]
This is a standard glued category for canonicity of dependent type
theory, and carries a category with families structure that we will
reuse from prior work for the (non-linear) universes,
$\Pi/\Sigma/\textrm{Id}$ \cite{coquand}. Concretely an object of
$\mathcal G$ is equivalent to a pair of a syntactic context $\Gamma$,
and a family $\hat \Gamma$ over (equivalence classes of) closing
substitutions for $\Gamma$.

Next we turn to interpreting the linear types and morphisms. For this
we define a functor $\gluedL : \gluedNL \to \textrm{MonCat}$ where
$\textrm{MonCat}$ is the category\footnote{in univalent foundations
this would be the category of \emph{strict} monoidal categories.} of
monoidal categories and lax monoidal functors.  We construct
$\gluedL(p : \hat \Gamma \to S(\cdot,\Gamma))$ as the following comma
category:

% https://q.uiver.app/#q=WzAsNixbMCwwLCJcXGdsdWVkTkwocCkiXSxbMCwyLCJMX1xcR2FtbWEiXSxbMSwyLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oUyhcXGNkb3QsXFxHYW1tYSkpIl0sWzIsMiwiXFxtYXRoY2FsIFAgXFxTaWdtYV4qL1xcdGV4dHJte0Rpc2N9KFxcaGF0XFxHYW1tYSkiXSxbMiwwLCJcXG1hdGhjYWwgUCBcXFNpZ21hXiovXFx0ZXh0cm17RGlzY30oXFxoYXRcXEdhbW1hKSJdLFsxLDEsIlxccm90YXRlYm94W29yaWdpbj1jXXstNDV9e1xcRG93bmFycm93fSJdLFs0LDMsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsNF0sWzAsMV0sWzEsMiwiTiIsMl0sWzIsMywiXFx0ZXh0cm17RGlzY30ocF4qKSIsMl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{\gluedNL(p)} \&\& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)} \\
	\& {\rotatebox[origin=c]{-45}{$\Downarrow$}} \\
	{L_\Gamma} \& {\mathcal P \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))} \& {\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}
	\arrow[Rightarrow, no head, from=1-3, to=3-3]
	\arrow[from=1-1, to=1-3]
	\arrow[from=1-1, to=3-1]
	\arrow["N"', from=3-1, to=3-2]
	\arrow["{\textrm{Disc}(p^*)}"', from=3-2, to=3-3]
\end{tikzcd}\]

Where $N : L_\Gamma \to {\mathcal P
  \Sigma^*/\textrm{Disc}(S(\cdot,\Gamma))}$ is the ``dependent nerve''
of the inclusion $\lceil\cdot\rceil : \Sigma^* \to L_\Gamma$ of
strings into linear contexts defined in indexed style as
\[ N(\Gamma \vdash A) \gamma w = \lceil w \rceil \vdash \cdot :  \simulsubst A \gamma \]
Then observe that $N$ and ${\textrm{Disc}(p^*)}$ are lax monoidal and
${\mathcal P \Sigma^*/\textrm{Disc}(\hat\Gamma)}$ is a topos so the
criteria for Hyland and Schalk's gluing construction apply, defining a
biclosed monoidal structure on $\gluedNL(p)$ as well as finite
(co-)products. Further, these constructions are all stable under
reindexing in $\gluedNL$, and so interpret the dependent linear typing
constructions for $I,\otimes,0,\oplus,\top,\amp,\lto,\tol$.

Concretely, an object of $\gluedNL(p)$ is equivalent to a pair of a
non-linear type $\Gamma \vdash A$ and a formal grammar $\hat A$
indexed by $\hat \Gamma$ with a projection function from $\hat A$ to
$N(A)$. We can think of this presheaf $\hat A$ as the presheaf of
canonical forms of $\simulsubst A \gamma$ for closing substitutions
$\gamma$ in contexts of the form $\lceil w \rceil$.

All that remains in this model is to define the semantics of the
non-standard connectives. We define them in an indexed style. First we
define the interpretation of a non-linear type $\Gamma \vdash X$ as a
family of sets $\hat X$ indexed by a closing substitution $\gamma :
S(\cdot,\Gamma)$, a semantic substitution $\hat \gamma : \hat
\Gamma(\gamma)$ and a closed term $\cdot \vdash e : \simulsubst A
\gamma$. Second, we define a linear type $\Gamma \vdash A$ as a family
$\hat A$ of grammars indexed by $\gamma,\hat\gamma$ as above along
with a morphism $\pi : \hat A \gamma \hat \gamma \to N(\simulsubst A
\gamma)$.

\begin{enumerate}
\item $\hat{c} \gamma \hat \gamma w = (w = c)$ and $\pi(*) = \lceil c \rceil = x_0:c \vdash x_0:c$
\item $\widehat{(\LinSigTy x X A)} \gamma \hat \gamma w = \Sigma(\cdot \vdash e : \simulsubst X \gamma)\Sigma(\hat x: \hat X \gamma \hat \gamma e) \hat{A}(\gamma,e)(\hat\gamma,\hat x) w$
\item $\sem{\LinPiTy x X A} \gamma w = \Sigma(x:\simulsubst X \gamma;\lceil w\rceil \vdash e:\simulsubst A \gamma)\Pi(\cdot \vdash e' : \simulsubst X \gamma)\Pi(\hat x:\hat{X}\gamma\hat\gamma x)\Sigma(\hat a : \hat A (\gamma,e')(\hat\gamma,\hat x) \times (\pi(\hat a) = \subst e {e'} x))$
  and $\pi(e,f) = \dlamb x.e$
\item $\widehat{G A} \gamma \hat \gamma e = \Sigma(\hat a: \hat{A} \gamma \hat\gamma \varepsilon) (p\gamma\hat\gamma \varepsilon(\hat a) = \mathsf{G}^{-1}e)$
\item $\sem{L_i}\gamma = \gluedNLUniv(\gamma)$
\end{enumerate}
